Labor Market Income and Substitution Effects Explain Contemporary Natural Selection

### 2020-12-18 why-natural-selection.Rmd
Compare effect sizes with and without childless. Be more tentative about any conclusions from comparing generations.

Selection effects are highly correlated across the two generations, and most
share the same sign (Appendix Figure \@ref(fig:plot-res-sibs)). Effect size
estimates are larger for the parents' generation. We treat this result cautiously,
because when we split respondents up by year of birth, we find few differences
in effect sizes between early- and late-born respondents, for either generation.
In other words, since estimated effect sizes change when we change the dependent
variable, but do not change over time within either dependent variable, this
may be due to remaining ascertainment bias within the sample.

@@

Figure \@ref(fig:plot-res-sibs) shows regressions of *number of siblings* on
polygenic scores. By definition, people who had no children cannot be included in
this data. For a cleaner comparison, we rerun regressions on the respondents'
generation (*number of children*) after excluding respondents with no
children.

@@

To check whether effect sizes were changing over time, we ran
regressions interacting polygenic scores with birth year, median split at 1950
("early born" versus "late born"). We use both *number of children* and
*number of siblings* as a dependent variable. We weight using
age/qualification cells, and further adjust for selection in the
parents' generation (see above).

Tables \@ref(tab:tbl-siblings-by-period) and
\@ref(tab:tbl-children-by-period) summarize the results. There is no
evidence for change in natural selection within the parents' generation. In the
respondents' generation, effect sizes were significantly larger in absolute size 
among the late-born for four polygenic scores: cognitive ability, EA2, EA3 and
extraversion. These changes are inconsistent with the intergenerational change, 
where effect sizes were if anything larger among the earlier, parents' generation.
Three scores showed significant changes in sign: alcohol use
(positive to negative); conscientiousness (negative to positive); and
type II diabetes (negative to positive). Overall, there is weak evidence for
change over time, and the clearest result is that the direction of selection is
broadly consistent.

>>>>

Selection effects are highly correlated across the two generations, and most
share the same sign (Appendix Figure \@ref(fig:plot-res-sibs)). Absolute effect
size estimates are larger for the parents' generation. We treat this result
cautiously, because when we split respondents up by year of birth, we find few
differences in effect sizes between early- and late-born respondents, for either
generation. In other words, since estimated effect sizes change across
"generations", but do not change over time within either generation, the change
may be due to remaining ascertainment bias within the sample or other effects.
In particular, effect sizes in both generations may depend on polygenic scores'
correlation with childlessness, and we cannot estimate this for the parents'
generation.

@@

```{r calc-res-sibs}
drake::loadd(res_wt_age_qual_weights)
drake::loadd(res_children_comparison)

cmp <- left_join(res_wt_age_qual_weights, res_children_comparison, by = "term")

cmp %>%
      mutate(
        term = fct_reorder(term, estimate.x, order_abs(1)),
        term = pretty_names(term)
      ) %>%
      ggplot(aes(y = term)) +
        geom_pointrange(aes(x = estimate.x, xmin = conf.low.x,
                            xmax = conf.high.x), color = "red", alpha = .8) +
        geom_pointrange(aes(x = estimate.y, xmin = conf.low.y,
                            xmax = conf.high.y), color = "blue", alpha = .8)

cor_inc_vs_excl_childless <- cor(cmp$estimate.x, cmp$estimate.y)
n_smaller <- sum(abs(cmp$estimate.y) < abs(cmp$estimate.x))

pct_diff <- abs(cmp$estimate.y)/abs(cmp$estimate.x) - 1
median_pct_diff <- median(pct_diff)

Figure \@ref(fig:plot-res-sibs) shows regressions of *number of siblings*, i.e.
parents' number of children, on polygenic scores. By definition, members of the
parents' generation who had no children cannot be included in this data. For a
clean comparison with the respondents' generation, we rerun regressions on
*number of children* excluding those with no children, and show
results in the figure.

Excluding childless people in the parents' generation could bias our estimates.
To learn about this, we compare effect sizes excluding and including childless
people in the *current* generation. The correlation
between the two sets of effect sizes is `r cor_inc_vs_excl_childless`. So, patterns across
different scores are broadly similar whether the childless are counted or not. 
Unfortunately, however, absolute effect sizes are smaller when the
childless are excluded, for `r n_smaller` out of 33 scores; the median
percentage change is `r percent(median_pct_diff)`. The fact that childless
people have such a strong effect on estimates makes it hard to compare absolute
effect sizes across generations. In particular, since the parents' generation has
a different distribution of numbers of children, childless people may have had 
more or less effect in that generation.

@@

As an alternative approach, , we run
regressions interacting polygenic scores with birth year, median split at 1950
("early born" versus "late born"). We use both *number of children* and
*number of siblings* as a dependent variable. We weight using
age/qualification cells, and further adjust for selection in the
parents' generation (see above).

Tables \@ref(tab:tbl-siblings-by-period) and
\@ref(tab:tbl-children-by-period) summarize the results. There is no
evidence for changes in selection effects within the parents' generation. In the
respondents' generation, effect sizes were significantly larger in absolute size 
among the later-born for four polygenic scores: cognitive ability, EA2, EA3 and
extraversion. These changes are inconsistent with the intergenerational change, 
where estimated effect sizes were larger among the earlier, parents' generation.
Three scores showed significant changes in sign: alcohol use
(positive to negative); conscientiousness (negative to positive); and
type II diabetes (negative to positive).

Overall, there is weak evidence for change over time. The clearest results are
that (a) the direction of selection, and (b) the pattern of relative effect sizes
across scores, are broadly consistent over time.
